---
title: "simulation_gp"
author: "Joonwon Lee"
date: "2023-10-17"
output: pdf_document
---

```{r libraries, include=FALSE}
library(fields)
library(GpGp)
library(MASS)
library(ggplot2)
```

# Define functions needed for experiments

1. exp_cov(theta,range,d): returns an exponential covariance matrix.   
2. mat_cov(theta,range,d): returns a matern covariance matrix.   
3. gen_exp_data(n_samples, theta, range): generates a finite sample of gaussian process using an exponential covariance matrix.  
4. gen_mat_data(n_samples, theta, range): generates a finite sample of gaussian process using a matern covariance matrix.   

```{r functions, results='hide'}
exp_cov = function(theta,range,d){

  cov_mat = theta * exp(-d/range)
  chol_matrix = chol(cov_mat) # returns L^T where L is lower triangular.
  output = list(cov_mat=cov_mat,chol=chol_matrix)
  return(output)
}
mat_cov = function(theta,range,d){

  cov_mat = theta * (1 + d/range) * exp(-d/range) 
  epsilon <- 1e-6  # Small positive constant
  c_regularized <- cov_mat + epsilon * diag(dim(cov_mat)[1])
  chol_matrix = chol(c_regularized) # returns L^T where L is lower triangular.
  output = list(cov_mat=cov_mat,chol=chol_matrix)
  return(output)
}

# Generate finite subset of Gaussian Process
gen_exp_data <- function(n_samples, theta, range){
  # Define the range of the square 
  x_min <- 0
  x_max <- 10
  y_min <- 0
  y_max <- 10  
  
  # Generate random x and y coordinates randomly
  x <- runif(n_samples, min = x_min, max = x_max)
  y <- runif(n_samples, min = y_min, max = y_max)
  # x <- seq(0,10, length.out = n)
  # y <- seq(0,10, length.out = n)
  
  # distance matrix
  d <- sqrt(outer(x,x,"-")^2 + outer(y,y,"-")^2 )  
  
  chol_matrix = exp_cov(theta,range,d)$chol
  z_list <- rnorm(n_samples)
  sim_data <- t(chol_matrix) %*% z_list
  # set location matrix and design matrix X.
  locs <- as.matrix(data.frame(x = x, y = y))
  X <- cbind(rep(1,n_samples), locs)
  output = list(sim_data = sim_data, locs = locs, X = X)
  return(output)
}

gen_mat_data <- function(n_samples, theta, range){
  # Define the range of the square 
  x_min <- 0
  x_max <- 10
  y_min <- 0
  y_max <- 10  
  x <- runif(n_samples, min = x_min, max = x_max)
  y <- runif(n_samples, min = y_min, max = y_max)

  d <- sqrt(outer(x,x,"-")^2 + outer(y,y,"-")^2 )  # Distance matrix 
  
  chol_matrix = mat_cov(theta,range,d)$chol
  z_list <- rnorm(n_samples)
  sim_data <- t(chol_matrix) %*% z_list # simulate data
  
  locs <- as.matrix(data.frame(x = x, y = y)) # location matrix
  X <- cbind(rep(1,n_samples), locs) # design matrix X
  
  output = list(sim_data = sim_data, locs = locs, X = X)
  return(output)
}
```

### covfun_name option in GpGp

We can fit the model using fit_model function in GpGp package by specifying covariance function as below:

matern15_scaledim:
$M(x,y) = \sigma^2 (1+|D^{-1}|x-y||)exp(-||D^{-1}x-y||),$

exponential_isotropic:
$M(x,y) = \sigma^2 exp(-||x-y||/\alpha).$ 

### Remark: options in fit_model():
                   
1. reorder=FALSE: maxmin ordering is not used unless nrow(locs) > 1e5. 
                       
2. m_seq: By default, a 10-neighbor approximaation is maximized , then a 30-neighbor approximation is maximized using 10 neighbor estimates as starting values. 


# GpGp experiments
1. Truth covariance function is exponential, fitting model is matern.
2. Truth covariance function is matern, fitting model is exponential.

## 1. Truth covariance function is exponential, fitting model is matern:
Recall that exponential covariance function is:
$M(x,y) = \sigma^2 \cdot exp(-||x-y||/\alpha).$

```{r experiment_1,cache = TRUE, results='hide'}
n_samples = seq(500,2000, 500 )
scale_par = matrix(0, nrow = length(n_samples), ncol= 2)
i=1
theta = 2
range = 2
for (n in n_samples){
  # Generate data using exponential_isotropic
  exp_data = gen_exp_data(n_samples=n, theta=theta, range=range)
  # fit the model using 'matern15_scaledim'
  # fixed_parms = c(2,3) means fix range_1 and range_2. It must be used with
  # start_parms.
  mod1 = fit_model(exp_data$sim_data, locs = exp_data$locs, X = exp_data$X ,
                  covfun_name = 'matern15_scaledim', fixed_parms = c(2,3), start_parms = c(3,2,2,0))
  
  # fit the model using the true covariance function
  mod2 = fit_model(exp_data$sim_data, locs = exp_data$locs, X = exp_data$X,
                  covfun_name = 'exponential_isotropic',
  fixed_parms = c(2), start_parms = c(3,2,0))
  scale_par[i,1] = mod1$covparms[1]
  scale_par[i,2] = mod2$covparms[1]
  if (i< length(n_samples)){i = i + 1}
  else {}
}

df1 = data.frame(n = n_samples , sp_mat = scale_par[,1], sp_mat2 = scale_par[,2])
gg <- ggplot() +
  geom_line(data = df1, aes(x= n, y=sp_mat, color="matern"), linewidth=1) +
  geom_line(data = df1, aes(x= n, y=sp_mat2, color="exponential"), linewidth=1) +
  geom_hline(yintercept = theta, color = "red", linetype = "dashed") +
  labs(title = "Data is generated using exponential cov", x = "n_samples", y = "theta estimates") +
  scale_color_manual(values = c("matern" = "blue", "exponential" = "green"))

print(gg)
```

## 2. Truth covariance function is matern, fitting model is exponential:
In this section, generate data using
$M(x,y) = \sigma^2 (1+||x-y||/\alpha)exp(-||D^{-1}x-y||/\alpha),$ and fit the model using exponential covariance function.

```{r experiment_2, cache = TRUE, results='hide'}
n_samples = seq(500,2000, 500 )
scale_par = matrix(0, nrow = length(n_samples), ncol= 2)
i=1
theta = 2
range = 2
for (n in n_samples){
  mat_data = gen_mat_data(n_samples=n, theta=theta, range=range)
  
  mod3 = fit_model(mat_data$sim_data, locs = mat_data$locs, X = mat_data$X ,
                  covfun_name = 'matern15_scaledim', fixed_parms = c(2,3), start_parms = c(3,2,2,0))
  
  mod4 = fit_model(mat_data$sim_data, locs = mat_data$locs, X = mat_data$X,
                  covfun_name = 'exponential_isotropic',
  fixed_parms = c(2), start_parms = c(3,2,0))
  scale_par[i,1] = mod3$covparms[1]
  scale_par[i,2] = mod4$covparms[1]
  if (i< length(n_samples)){i = i + 1}
}

df1 = data.frame(n = n_samples , sp_mat = scale_par[,1], sp_mat2 = scale_par[,2])
gg <- ggplot() +
  geom_line(data = df1, aes(x= n, y=sp_mat, color="matern"), size=1) +
  geom_line(data = df1, aes(x= n, y=sp_mat2, color="exponential"), size=1) +
  geom_hline(yintercept = theta, color = "red", linetype = "dashed") +
  labs(title = "Data is generated using matern cov", x = "n_samples", y = "theta estimates") +
  scale_color_manual(values = c("matern" = "blue", "exponential" = "green"))

print(gg)

```

# Sanity check using MLE
Given a sample generated from exponential covariance matrix, we can try fitting scale parameter using MLE of covariance matrix:

```{r sanity_check}
# Generate 10000 observations of multivariate normal
# having exponential covariance function
n2 = 1000
x_min <- 0
x_max <- 10
y_min <- 0
y_max <- 10  
x <- runif(n2, min = x_min, max = x_max)
y <- runif(n2, min = y_min, max = y_max)
d <- sqrt(outer(x,x,"-")^2 + outer(y,y,"-")^2 ) 
theta = 2; range = 2
cov_mat <- theta * exp(-d/range)
data <- mvrnorm(n = n2, mu = rep(0,n2), Sigma = cov_mat)

# mle of covariance matrix
mle_cov = 1/n2 * t(data) %*% data

# fit model using matern15_scaledim: sigma^2 * (1+d) * exp(-d/range)

sigma_sq = diag(mle_cov %*% solve( (1+d/range) * exp(-d/range) ))

# fit model using true exponential: sigma^2 * exp(-d/range)

sigma_sq2 = diag(mle_cov %*% solve(exp(-d/range) ))

cat("Given ", n2, "samples with true [theta, range] =", matrix(c(2,2),nrow=1),
 ",\n\nestimate of (theta, range ) by fitting with matern15_isotropic are :", mean(sigma_sq),
"\n\nestimate of (theta, range ) by fitting with true exponential_isotropic are :", mean(sigma_sq2) )
```
# Predictions

Given new locations, we can predict new y values using predictions() in GpGp package. Then we can fit a new model using predicted values. 

```{r predictions, cache = TRUE, results='hide'}
n_new = 1000
x_min <- 0
x_max <- 10
y_min <- 0
y_max <- 10  
x_new <- runif(n_new, min = x_min, max = x_max)
y_new <- runif(n_new, min = y_min, max = y_max)
locs_new <- as.matrix(data.frame(x = x_new, y = y_new)) # location matrix
X_new <- cbind(rep(1,n_new), locs_new) # design matrix X

pred = predictions(
fit = mod2,
locs_pred = locs_new,
X_pred  = X_new,
y_obs = mod2$y,
locs_obs = mod1$locs,
X_obs = mod2$X,
beta = mod2$betahat,
covparms = mod2$covparms,
covfun_name = mod2$covfun_name,
m = 60,
reorder = TRUE,
st_scale = NULL
)

mod5 = fit_model(pred, locs = locs_new, X = X_new,
                  covfun_name = 'exponential_isotropic',
  fixed_parms = c(2), start_parms = c(3,2,0))

cat("expected estimate of theta given new locations: ", mod5$covparms[1])
```